\section{Related work}

This project builds on much existing material by being both a modern rewrite of an existing tool and a case study about the viability of a number of technologies for porting complex desktop applications to the browser.

It naturally is closely related to the Nand to Tetris~\cite{n2tweb} course. All the high-level functionality was designed and specified by Nisan and Schocken.
Since the goal of this project is to develop an alternative emulator implementation, care was taken to follow the specifications described in the companion book, specifically the sections 4.2, 7.2, 8.2 and 9.2.7~\cite{nisan2005}.
In addition, many of the integration tests are direct ports of test cases from the Nand to Tetris course.

% \begin{itemize}
%   \item VM Bytecode design und high level Funktionalität
%   \item CPU Assembly design und high level Funktionalität
%   \item TST design und high level Funktionalität
%   \item Viele Tests für Emulatoren aus N2T Projekten
% \end{itemize}

\subsection{Existing emulator implementations}
The main implementation of a Nand to Tetris Emulator is, of course, the official one, which can be downloaded directly from the course website and contains every necessary tool to complete all projects~\cite[Software]{n2tweb}.
\label{pynand}
That being said, there are other people who have tried to provide an alternative emulator experience.
One of these is pynand, which aims to provide a better experience by eliminating the need to install Java and by making the interface less ``clunky''~\cite{pynand}.
However, the developer made several decisions that limit its usefulness.
It does little to improve the performance of the emulator and also severely limits the programs that can be run.
This limitation of possible programs is a direct consequence of its architecture, which translates the VM bytecode~[\ref{hack-bytecode}] into machine code for the hack architecture~[\ref{hack-architecture}] instead of interpreting it directly.
At first glance this may seem reasonable, since it is closer to a theoretical implementation of the system in hardware, but it severely limits the possibilities of the VM emulator, since each VM instruction corresponds to multiple assembly instructions.
This causes more complex VM programs to quickly overflow the available memory for instructions, limiting the emulator to relatively short programs.
Since the goal of this project was to partially replace the official VM emulator implementation, which allows for larger programs than should theoretically be possible, this approach is not an option here.
It also doesn't provide a user interface outside the display, which makes debugging complex programs difficult.
Besides, the user still has to install software on their computer, just Python instead of Java.

% \begin{itemize}
%   \item https://github.com/itoshkov/nand2tetris-emu
%   \item https://github.com/mossprescott/pynand
% \end{itemize}

% \subsection{Emulators in WebAssembly}
In recent years, there have been multiple projects that implemented other emulators in WebAssembly. Notably, an emulator for Intel's 8086 microprocessor~\cite{9824078}, also written in Rust and using a ReactJS front-end. This already demonstrates the viability of these technologies for creating web-based emulators.

% \begin{itemize}
%   \item https://ieeexplore.ieee.org/abstract/document/9824078
%   \item https://wasm4.org/
% \end{itemize}
\subsection{Dependencies}
\label{rust-deps}
One of the main reasons why Rust was chosen as the language for this project is its strong ecosystem, especially in regard to WebAssembly~[\ref{rust-vs-other-wasm}].
Therefore, it is not surprising that several other Rust projects were used here to be both more efficient and to produce a more stable and reliable final program.
At the same time, it is also important to use as few dependencies as possible, as any dependency could become a future liability if that library is ever abandoned or compromised by malicious actors.
For this reason, each library used in this project has a clear purpose and provides enough value to justify its inclusion. Libraries are often referred to as crates in Rust, so these terms are used interchangeably throughout the text.

This section summarizes the rationale for including each crate.
Rust does not currently allow static constants that require runtime code for initialization.
This is however a very useful feature, for example to use hashmaps for defining lookup tables.
The \verb+lazy_static+ crate is one of the most used libraries in the Rust ecosystem~\cite[Downloads all time]{lazystatic}, because it provides exactly this functionality in the form of a single macro.
Among other things, it is used for the lookup table in the keyboard handler that maps special symbols like the escape key to the corresponding key codes in the emulator.

Regular expressions are a very handy tool for simple parsing tasks. They are used to parse some tokens in the test scripts~[\ref{test-scripts}] for the emulator. Rust does not provide regular expressions in its standard library, so the \verb+regex+ crate was used.

\label{web-sys}
The two most important libraries when it comes to the value of Rust for this project are \verb+wasm-bindgen+ and \verb+web-sys+.
While the former is responsible for creating bindings between Rust and JavaScript so that the two can call each other as if they were the same language, the latter provides implementations for common JavaScript APIs in Rust.
It builds directly on top of \verb+wasm-bindgen+ to provide most of the functionality that users would normally have to manually write bindings for.
These two libraries are further complemented by \verb+console_error_panic_hook+ and \verb+wasm-pack+.
The former simply prints the stack trace of any unrecoverable Rust error~[\ref{error-handling}] to the JavaScript console.
\verb+Wasm-pack+, on the other hand, is a build tool that simplifies working with Rust in the context of Wasm by providing a command line utility which automatically builds Rust applications and packages them into Wasm modules.
This tool is only used at compile time and is therefore only needed to compile the application, not to run it.

The above dependencies are all that is needed for the web version of the emulator. If the native version is used instead, two different dependencies are used instead of the Wasm-focused dependencies.
The first is \verb+SDL2+, a library for building native desktop applications and games.
It is included only if the application was compiled with the desktop flag enabled~[\ref{conditional-compilation}].
Finally, there is \verb+clap+, an easy-to-use parsing library for command line arguments.

% \begin{itemize}
% \item lazy\_static (hack um rust weniger nervig zu machen)
% \item regex
% \item wasm-bindgen (rust code für JS zugänglich machen)
% \item web-sys (js stdlib in Rust nutzen)
% \item console\_error\_panic\_hook (rust panics zu JS exeptions)
% \item sdl2 (native UI (eigentlich nur zum Testen))
% \item clap (CLI parsing)
% \item wasm-pack (rust -> wasm Kompilierung einfacher machen)
% \end{itemize}

% \subsubsection{JavaScript Dependencies}
ReactJS is the only JavaScript dependency for the web-based user interface.
Since most of the logic is implemented in Rust, the code required for the user interface is mainly focused on displaying the internal state of the emulator.
There are no complex UI components or procedures that would justify the use of separate libraries.
The entire front-end simply uses React hooks for all state handling.
In addition, the interface was intentionally kept minimalist to match the emulator's black-and-white display, so no external CSS libraries are required.

The appendix also contains a table with every dependency used in the project~[\ref{table:dependencies}].
